Emergent dual scaling of riverine biodiversity

### 2021-06-21 manuscript_pnas.Rmd
edits: manscript and SI

  # read theory file
  dat_sim <- read_csv(here::here('code_theory/result/result_sim2020-11-10.csv'))

@@

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity [@turnerLandscapeEcologyTheory2015; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few [@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike [@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms [@grantLivingBranchesPopulation2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure [@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @brownDendriticNetworkStructure2010], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems [@rinaldoRiverNetworksEcological2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches [@carraraDendriticConnectivityControls2012]. However, most empirical research has explored the consequences of branching complexity for local community structure [@erosLandscapeEcologyRivers2019] or has relied solely on theoretical arguments with limited replications of branching architecture [@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity patterns at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services [@reidEmergingThreatsPersistent2019].

@@

We developed a theoretical framework synthesizing the influences of ecosystem size and complexity on biodiversity patterns in branching ecosystems. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit [@yeakelSynchronisationStabilityRiver2014; @teruiMetapopulationStabilityBranching2018] (**Figure 1A and B**). Two factors determine the mean environmental condition at each habitat patch $\mu_z$: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution with a mean of zero and a standard deviation of $\sigma_h$ and propagate downstream. We modeled the downstream propagation as a spatial autoregressive process, in which local noise is added to the adjacent upstream environment as $\mu_{z,k} \sim Normal(\mu_{z,k+1}, \sigma_{l}^2)$ ($k$ is the network distance from the outlet patch; $k = 1$ at the outlet) . The environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Materials and Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity [@mcguireNetworkAnalysisReveals2014].

@@

Our simulation predicted the emergence of dual scaling law under a realistic landscape scenario, where the environmental variation at headwaters $\sigma_h$ is greater than the degree of local environmental noise $\sigma_l$ (**Figure 1A**). Ecosystem size and complexity both increased $\gamma$ diversity (**Figures 2A and 3A**) with a signature of power law (i.e., linear in a log-log scale), and these relationships were consistent across ecological scenarios (compare panels in **Figures 2A and 3A**). Importantly, the impact of ecosystem complexity was comparable to that of ecosystem size. Hence, regardless of ecological scenarios, ecosystem size and complexity are likely to be equally important in scaling $\gamma$ diversity. Our results contrast with predictions from neutral models, in which the enhanced connectivity of highly branched networks reduced $\gamma$ diversity by facilitating the further expansion of an already widespread species [@muneepeerakulEvolvingBiodiversityPatterns2019]. Non-neutral components of our model should therefore have played a central role in generating the predicted patterns. Indeed, the robustness of the scaling law stemmed from the convergence of niche and dispersal assembly as described below.

Dispersal regulated community assembly processes that converged to the robust scaling relationships of $\gamma$ diversity (compare left and right columns in **Figures 2A and 3A**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This result reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@leiboldMetacommunityConceptFramework2004; @thompsonProcessbasedMetacommunityFramework2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@leiboldMetacommunityConceptFramework2004; @economoSpeciesDiversityNeutral2008; @thompsonProcessbasedMetacommunityFramework2020]. Therefore, ecosystem size and complexity can increase $\gamma$ diversity via enhanced species sorting or mass effects, and which takes primacy is dependent on the degree of dispersal limitation. These patterns were qualitatively similar across different levels of dispersal probabilities (**Figures S5 and S12**).

@@

@@

Recent advancement of spatial theories has sparked discussion of how the complex spatial structure of ecosystems drives macroecological patterns of diverse organisms. Several studies have explicitly considered the physical architecture of branching networks [@rinaldoRiverNetworksEcological2018] and have succeeded in reproducing some of the core biodiversity patterns (e.g., spatial distribution of local species richness) at coarse spatial resolutions (\~5000 km^2^) [@muneepeerakulNeutralMetacommunityModels2008a]. Yet, we have overlooked one of the fundamental rules in complex ecosystems, i.e., the dual scaling law. Here, the unified framework of ecosystem size and complexity provided strong evidence for the dual scaling law. Notably, this pattern has emerged irrespective of ecological contexts or geographic regions, emphasizing the generality of our findings. The dual scaling law of biodiversity may be an overlooked yet pervasive property of branching ecosystems.

@@

While the prevailing evidence supports the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in community assembly processes [@chaseStochasticCommunityAssembly2010]. Recognizing the dual scaling of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Michio Fukushima and Hirokazu Urabe for providing information on the Hokkaido Freshwater Fish Database HFish and the fish monitoring data at protected watersheds in Hokkaido. We thank Jacques Finlay for his constructive comments on the earlier version of the manuscript. This material is based upon work supported by the National Science Foundation through a Water Sustainability and Climate Program Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through the Division of Environmental Biology (DEB 2015634), Environmental Protection Agency through a Water Quality Benefits grant (EPA-G2015-STAR-A1), and the University of North Carolina at Greensboro through a start-up fund.

>>>>

  # read theory file
  dat_sim <- read_csv(here::here('code_theory/result/result_sim2021-05-28.csv'))

@@

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity [@turnerLandscapeEcologyTheory2015; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few [@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike [@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms [@grantLivingBranchesPopulation2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure [@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @brownDendriticNetworkStructure2010], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems [@rinaldoRiverNetworksEcological2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches [@carraraDendriticConnectivityControls2012]. However, most research has explored the consequences of branching complexity for local community structure [@erosLandscapeEcologyRivers2019] or has relied solely on theoretical arguments with limited replications of branching architecture [@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity patterns at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services [@reidEmergingThreatsPersistent2019].

@@

We developed a theoretical framework synthesizing the influences of ecosystem size and complexity on biodiversity patterns in branching ecosystems. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit [@yeakelSynchronisationStabilityRiver2014; @teruiMetapopulationStabilityBranching2018] (**Figure 1A and B**). In this network, the number of habitat patches $N_p$ (a theoretical proxy for ecosystem size $A$) and branching probability $P_b$ (ecosystem complexity) shape the spatial structure (**Materials and Methods**). We introduced two factors determining the temporal mean of environmental condition at each habitat patch $\mu_z$: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution with a mean of zero and a standard deviation of $\sigma_h$ and propagate downstream. We modeled the downstream propagation as a spatial autoregressive process, in which local noise is added to the adjacent upstream environment as $\mu_{z,k} \sim Normal(\mu_{z,k+1}, \sigma_{l}^2)$ ($k$ is the network distance from the outlet patch; $k = 1$ at the outlet) . The environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Materials and Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity [@mcguireNetworkAnalysisReveals2014].

@@

Our simulation predicted the emergence of dual scaling law under a realistic landscape scenario, where the environmental variation at headwaters $\sigma_h$ is greater than the degree of local environmental noise $\sigma_l$ (**Figure 1A**). Ecosystem size and complexity both increased $\gamma$ diversity (**Figures 2A and 3A**) with a signature of power law (i.e., linear in a log-log scale), and these relationships were consistent across ecological scenarios (compare panels in **Figures 2A and 3A**). Importantly, the impact of ecosystem complexity was comparable to that of ecosystem size. Hence, regardless of ecological scenarios, ecosystem size and complexity are likely to be equally important in scaling $\gamma$ diversity. Our results contrast with predictions from neutral models, in which the enhanced connectivity of highly branched networks reduced $\gamma$ diversity by facilitating the further expansion of an already widespread species [@muneepeerakulEvolvingBiodiversityPatterns2019]. Non-neutral components of our model should therefore have played a central role in generating the predicted patterns. Indeed, the robustness of the scaling law stemmed from the convergence of niche and dispersal assembly processes as described below.

Dispersal strength regulated the relative importance of ecological mechanisms that underlie the scaling relationships of $\gamma$ diversity (compare left and right columns in **Figures 2A and 3A**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This result reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@leiboldMetacommunityConceptFramework2004; @thompsonProcessbasedMetacommunityFramework2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@leiboldMetacommunityConceptFramework2004; @economoSpeciesDiversityNeutral2008; @thompsonProcessbasedMetacommunityFramework2020]. Therefore, ecosystem size and complexity can increase $\gamma$ diversity via enhanced species sorting (niche assembly) or mass effects (dispersal assembly), and which takes primacy is dependent on the degree of dispersal limitation. These patterns were qualitatively similar across different levels of dispersal probabilities (**Figures S5 and S12**).

@@

While our empirical results showed an excellent match with the predictions of our niche-based model (**Figures 2 and 3**), we must note that a neutral model developed by Muneepeerakul et al. [@muneepeerakulNeutralMetacommunityModels2008a] was also capable of reproducing fish diversity patterns in the Mississippi-Missouri River Basin. This apparent contradiction is perhaps not a surprise when considering the difference in the spatial resolution of interest. Their neutral model used sub-basins (ca. 5000 km^2^ in the watershed area) as a unit of local communities (metacommunities in our study) and modeled variation in habitat capacity and dispersal among those units. At this coarse resolution, stochastic immigration processes coupled with random speciation, a core of the Unified Neutral Theory of Biodiversity [@hubbellUnifiedNeutralTheory2001], are critical to forming biogeographic patterns [@carvajal-quinteroDrainageNetworkPosition2019]. Meanwhile, we defined local communities at much finer resolutions (several tens to hundred meters) at which species' niche should interact with environments to determine community structure [@jacksonWhatControlsWho2001]. It is therefore necessary for our theoretical model to assume niche-based processes. This interpretation is supported by the fact that a neutral model could not reproduce the positive complexity-diversity relationship observed in our empirical data [@muneepeerakulEvolvingBiodiversityPatterns2019]. As such, our research is uniquely significant in showing the qualitative agreement of theoretical and empirical results at this level of biological organization.

@@

The past two decades have witnessed efforts to illuminate the role of spatial ecosystem structure in shaping biodiversity patterns [@rinaldoRiverNetworksEcological2018; @grantLivingBranchesPopulation2007]. Yet, current schemes have overlooked one of the fundamental rules in branching ecosystems, i.e., the dual scaling law. Here, our unified framework unveiled the surprising consistency of dual scaling across ecological contexts or geographic regions, suggesting the universality of our findings. This discovery brings a paradigm shift in macroecological science.

@@

While prevailing evidence supports the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in community assembly processes [@chaseStochasticCommunityAssembly2010]. Recognizing the dual scaling of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Michio Fukushima and Hirokazu Urabe for providing information on the Hokkaido Freshwater Fish Database HFish and the fish monitoring data at protected watersheds in Hokkaido. We thank Jacques Finlay for his constructive comments on the earlier version of the manuscript. This material is based upon work supported by the National Science Foundation through the Water Sustainability and Climate Program Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through the Division of Environmental Biology (DEB 2015634), Environmental Protection Agency through the Water Quality Benefits grant (EPA-G2015-STAR-A1), the University of North Carolina at Greensboro through a start-up fund, and the Japan Society for the Promotion of Science through the KAKENHI grant (20H03010).
